---
title: "CombinedAnalysis"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read the count matrix and experiment data

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
#read counts data 
cts_prime <- as.matrix(read.csv("sorted.csv", header=T, sep=","))

#add a column with the name of the samples
cts<-cts_prime[,c("SRR1027171","SRR1027173","SRR1027174","SRR1027175","SRR1027176","SRR1027177","SRR1027178","SRR1027179","SRR1027180",
"SRR1027181","SRR1027182","SRR1027183","SRR1027184","SRR1027185","SRR1027186","SRR1027187","SRR1027188", "SRR1027189","SRR1027190")]

row.names(cts) <- cts_prime[,1] #append the gene names to the samples

cts<-apply(cts, c(1,2), as.integer) #change the gene names from chars to ints

cts<-cts[complete.cases(cts), ] #remove rows with empty data cells

coldata<-read.csv("coldata.csv",
                  header = T,
                  row.names=1,
                  stringsAsFactors=T)#read the experiment info
```

Creates the DESeqDataSet based on the different cancer types and the control group 

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
library("magrittr")
levels(coldata$cancer_type) #check if the control group is first
levels(coldata$cancer_type) <- c("NBS","HER2","Non-TBNC","TNBC")
coldata$cancer_type <- relevel(coldata$cancer_type, "NBS") #place the control group first 
levels(coldata$cancer_type) #validate the changes

#create the des object by looking at the different groups of cancer and control group
library("DESeq2")
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ cancer_type)

#remove rows with low insignificant counts
dds <- dds[rowSums(counts(dds)) > 1,]


# DESeq2 offers two transformations for count data that stabilize the variance across the mean:     
# -The variance stabilizing transformation (VST) for negative binomial data with a dispersion-mean trend implemented in the vst function
# -The regularized-logarithm transformation implemented in rlog() function.


rld <- rlog(dds,blind = FALSE)
#head(assay(rld), 3)

vsd <- vst(dds,blind = FALSE)
#head(assay(vsd), 3)

# In the above function calls, we specified blind = FALSE, which means that differences between cell lines and treatment (the variables in the design) will not contribute to the expected variance-mean trend of the experiment. The experimental design is not used directly in the transformation, only in estimating the global amount of variability in the counts. For a fully unsupervised transformation, one can set blind = TRUE ( the default).

```

Figure 1: Create a heatmap of the samples

```{r eval=TRUE, echo=TRUE, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}

sampleDists <- dist(t(assay(rld)))

library ("pheatmap")
library ("RColorBrewer")

sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <- paste( rld$cancer_type, sep = " - " )
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

library(ComplexHeatmap)
my_hclust_gene <- hclust(sampleDists, method = "complete")
 
# install if necessary
#install.packages("dendextend")
 
# load package
library(dendextend)
 
as.dendrogram(my_hclust_gene) %>%
  plot(horiz = TRUE)
my_gene_col <- cutree(tree = as.dendrogram(my_hclust_gene), k = 4)
my_gene_col <- data.frame(cluster = ifelse(test = my_gene_col == 1, yes = "cluster 1", no = 
            ifelse(test = my_gene_col == 2, yes = "cluster 2", no =
            ifelse(test = my_gene_col == 3, yes = "cluster 3", no  = "cluster 4"))))
#head(my_gene_col)
#Heatmap of sample-to-sample distances using the regularized-logarithm transformed values
pheatmap(sampleDistMatrix, annotation_row = my_gene_col, col = colors)
```

Figure 2: Run PCA analysis

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
#Running PCA on the samples 
plotPCA(vsd, intgroup=c("cancer_type"))
```

Figure 3: Creates a heatmap of the top 20 up regulated genes in and cluster by value similarities 

```{r eval=TRUE, echo=TRUE, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
# In the sample distance heatmap made previously, the dendrogram at the side shows us a hierarchical clustering of the samples. Such a clustering can also be performed for the genes. Since the clustering is only relevant for genes that actually carry a signal, one usually would only cluster a subset of the most highly variable genes. Here, for demonstration, let us select the 20 genes with the highest variance across samples.

#rowVars: Variance estimates for each row in a matrix
topVarGenes <- head(order(rowVars(assay(vsd)), decreasing = TRUE), 20)

mat  <- assay(vsd)[topVarGenes, ]
mat<-mat-rowMeans(mat)
anno <- as.data.frame(colData(vsd)[, c("cancer_type")])
rownames(anno) <- colnames(mat)
pheatmap(mat, annotation_col = anno)

#Treatment status and cell line information are shown with colored bars at the top of the heatmap. Blocks of genes that covary across patients.
```

Creates the DESeqDataSet based on the different of normal breast sample and cancered breast 

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
library("magrittr")
levels(coldata$condition) #check if the control group is first
coldata$condition <- relevel(coldata$condition, "normal") #place the control group first 
levels(coldata$condition)#validate the changes

dds2 <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ condition)

#remove rows with low insignificant counts
dds2 <- dds2[rowSums(counts(dds2)) > 1,]

dds_analysis <- DESeq(dds2)
```

Lets look at the interesting counts (with low pvalue and high absolute value of log2foldchange)

```{r eval=TRUE, echo=TRUE, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}

library("EnhancedVolcano")
dds <- DESeq(dds2, betaPrior=FALSE)
res <- results(dds, contrast = c('condition','cancer', 'normal'))
res <- lfcShrink(dds,type = 'normal', contrast = c('condition','cancer', 'normal'),res=res) 
```


```{r eval=TRUE, echo=TRUE, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
EnhancedVolcano(res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'pvalue')
```

Lets keep only the significant counts by choosing p-value < 0.05 and log2foldchange > 1 or log2foldchange < -1

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
#Calling results without any arguments will extract the estimated log2 fold changes and p values for the last variable in the design formula (design = ~ condition)

#If we lower the false discovery rate threshold, we should also inform the results() function about it, so that the function can use this threshold for the optimal independent filtering that it performs:
# If we want to raise the log2 fold change threshold, so that we test for genes that show more substantial changes due to treatment, we simply supply a value on the log2 scale. For example, by specifying lfcThreshold = 1, we test for genes that show significant effects of treatment on gene counts more than doubling or less than halving, because 2^1=2.

res.05.LFC1 <- results(dds_analysis, alpha = 0.05, lfcThreshold=1)
table(res.05.LFC1$padj < 0.05)

#As res is a data frame that carries also a metadata with information on the meaning of the columns:
summary(res.05.LFC1)
```

Running combined analysis between normal and cancer:

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
#We subset the results table to the genes with FDR<0.1 and then sort it by the log2 fold change estimate to get the significant genes with the strongest down-regulation:
resSig <- subset(res.05.LFC1)
head(resSig[ order(resSig$log2FoldChange), ])

#The genes with the strongest up-regulation due to the cancer:
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```

Adding annotation to the gene names:

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
pacman::p_load("AnnotationDbi")
#fixes a bug in the communication between the latest release of the RSQLite package and RStudio. Without it the org.Hs.eg.db package can't be loaded.
options(connectionObserver = NULL)
library(org.Hs.eg.db)

columns(org.Hs.eg.db)

#The genes with the strongest up-regulation due to treatment
ens.str <- (rownames(res))
res.05.LFC1$ensembl <- mapIds(org.Hs.eg.db,
                     keys=ens.str,
                     column="ENSEMBL",
                     keytype="SYMBOL",
                     multiVals="first")
res.05.LFC1$entrez <- mapIds(org.Hs.eg.db,
                     keys=ens.str,
                     column="ENTREZID",
                     keytype="SYMBOL",
                     multiVals="first")

head(res.05.LFC1)
```

Exporting the results:

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
resOrdered <- res.05.LFC1[order(res.05.LFC1$pvalue),]
head(resOrdered)
resOrderedDF <- as.data.frame(resOrdered)[1:100, ]
write.csv(resOrderedDF, file = "resultsCombinedNormalVs.Cancer.csv")

```

Running separated analysis between normal and HER2:

```{r}
#read counts data 
cts_prime <- as.matrix(read.csv("sortedHER2.csv", header=T, sep=","))
#add a column with the name of the samples
cts<-cts_prime[,c("SRR1027183","SRR1027184","SRR1027185","SRR1027186","SRR1027187","SRR1027188", "SRR1027189","SRR1027190")]
#append the gene names to the samples
row.names(cts) <- cts_prime[,1]
#change the gene names from chars to integers
cts<-apply(cts, c(1,2), as.integer)
#remove rows with empty data cells
cts<-cts[complete.cases(cts), ]
#read the experiment info
coldata<-read.csv("coldataHER2.csv",
                  header = T,
                  row.names=1,
                  stringsAsFactors=T)

library("magrittr")

#------------------------for normal vs HER2
levels(coldata$cancer_type) #check if the control group is first
coldata$cancer_type <- relevel(coldata$cancer_type, "NBS", "TNBC","Non-TNBC","HER2") 
levels(coldata$cancer_type) #validate the changes
library("DESeq2")
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ cancer_type)
dds <- dds[rowSums(counts(dds)) > 1,]

dds_analysis <- DESeq(dds)
res.05.LFC1 <- results(dds_analysis, alpha = 0.05, lfcThreshold=1)
#table(res.05.LFC1$padj < 0.05)
summary(res.05.LFC1)
resSig <- subset(res.05.LFC1) 
head(resSig[ order(resSig$log2FoldChange), ]) #down-regulation
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ]) #up-regulation
```

Exporting the results:

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
resOrdered <- res.05.LFC1[order(res.05.LFC1$pvalue),]
head(resOrdered)
resOrderedDF <- as.data.frame(resOrdered)[1:100, ]
write.csv(resOrderedDF, file = "resultsNBSVs.HER2.csv")

```

Running separated analysis between normal and Non-TNBC:

```{r}
#read counts data 
cts_prime <- as.matrix(read.csv("sortedNon-TNBC.csv", header=T, sep=","))
#add a column with the name of the samples
cts<-cts_prime[,c("SRR1027177","SRR1027178","SRR1027179","SRR1027180",
"SRR1027181","SRR1027182","SRR1027188", "SRR1027189","SRR1027190")]
#append the gene names to the samples
row.names(cts) <- cts_prime[,1]
#change the gene names from chars to integers
cts<-apply(cts, c(1,2), as.integer)
#remove rows with empty data cells
cts<-cts[complete.cases(cts), ]
#read the experiment info
coldata<-read.csv("coldataNon-TNBC.csv",
                  header = T,
                  row.names=1,
                  stringsAsFactors=T)

library("magrittr")

#------------------------for normal vs Non-TNBC
levels(coldata$cancer_type) #check if the control group is first
coldata$cancer_type <- relevel(coldata$cancer_type, "NBS", "TNBC","HER2","Non-TNBC") 
levels(coldata$cancer_type) #validate the changes
library("DESeq2")
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ cancer_type)

dds <- dds[rowSums(counts(dds)) > 1,]

dds_analysis <- DESeq(dds)
res.05.LFC1 <- results(dds_analysis, alpha = 0.05, lfcThreshold=1)
#table(res.05.LFC1$padj < 0.05)
summary(res.05.LFC1)
resSig <- subset(res.05.LFC1) 
head(resSig[ order(resSig$log2FoldChange), ]) #down-regulation
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ]) #up-regulation

```

Exporting the results:

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
resOrdered <- res.05.LFC1[order(res.05.LFC1$pvalue),]
head(resOrdered)
resOrderedDF <- as.data.frame(resOrdered)[1:100, ]
write.csv(resOrderedDF, file = "resultsNBSVs.Non-TNBC.csv")

```

Running separated analysis between normal and TNBC:

```{r}
#read counts data 
cts_prime <- as.matrix(read.csv("sortedTNBC.csv", header=T, sep=","))
#add a column with the name of the samples
cts<-cts_prime[,c("SRR1027171","SRR1027173","SRR1027174","SRR1027175","SRR1027176","SRR1027188", "SRR1027189","SRR1027190")]
#append the gene names to the samples
row.names(cts) <- cts_prime[,1]
#change the gene names from chars to integers
cts<-apply(cts, c(1,2), as.integer)
#remove rows with empty data cells
cts<-cts[complete.cases(cts), ]
#read the experiment info
coldata<-read.csv("coldataTNBC.csv",
                  header = T,
                  row.names=1,
                  stringsAsFactors=T)

library("magrittr")

#------------------------for normal vs TNBC
levels(coldata$cancer_type) #check if the control group is first
coldata$cancer_type <- relevel(coldata$cancer_type, "NBS", "Non-TNBC","HER2","TNBC") 
levels(coldata$cancer_type) #validate the changes
library("DESeq2")
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ cancer_type)
dds <- dds[rowSums(counts(dds)) > 1,]

dds_analysis <- DESeq(dds)
res.05.LFC1 <- results(dds_analysis, alpha = 0.05, lfcThreshold=1)
#table(res.05.LFC1$padj < 0.05)
summary(res.05.LFC1)
resSig <- subset(res.05.LFC1) 
head(resSig[ order(resSig$log2FoldChange), ]) #down-regulation
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ]) #up-regulation

```

Exporting the results:

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
resOrdered <- res.05.LFC1[order(res.05.LFC1$pvalue),]
head(resOrdered)
resOrderedDF <- as.data.frame(resOrdered)[1:100, ]
write.csv(resOrderedDF, file = "resultsNBSVs.TNBC.csv")

```
